From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)

### 2025-04-15 analysis/paper/paper.qmd
Update institution

---
title: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
author:
  - Joana Belmiro:
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      orcid: https://orcid.org/0000-0003-4951-2907
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Nuno Bicho:
      orcid: https://orcid.org/0000-0001-9655-0549
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute:
        - Institución Milá y Fontanals de Investigación en Humanidades, Consejo Superior de Investigaciones Científicas, Barcelona, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  The Upper Palaeolithic (UP) of westernmost Europe was marked by technological and cultural transformations and abrupt climatic shifts. The Iberian Peninsula, particularly southwestern Iberia, served as a refugia and key eco-cultural niche, making it a key region for UP studies. Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time. This study examines lithic raw material procurement and technological organisation at Vale Boi, focusing on the differences between local and non-local cherts. We analysed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published techno-typological and raw material macroscopic and petrographic data. Our results indicate that non-local cherts were intensively reduced. The presence of varied techno-typological classes suggests that non-local cherts were not exclusively transported as finished tools, instead highlighting their role as versatile, transportable raw material volumes. Gravettian occupations exhibited greater reliance on non-local cherts, suggesting short-term occupations and higher mobility, while Proto-Solutrean and Solutrean occupations showed increased dependence on local materials, suggesting long-term occupations. These findings expand our understanding of UP technological adaptations in southwestern Iberia, emphasizing the complex role of lithic resources in mobility, settlement, and social networks.

>>>>

---
title: "From Stone to Tool: How Raw Materials Influenced Upper Paleolithic Technology in Southwestern Iberia (Vale Boi)"
author:
  - Joana Belmiro:
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      orcid: https://orcid.org/0000-0003-4951-2907
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Nuno Bicho:
      orcid: https://orcid.org/0000-0001-9655-0549
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
  - Xavier Terradas:
      orcid: https://orcid.org/0000-0002-8000-5607
      institute:
        - Spanish National Research Council - Institute Milá y Fontanals of research on Humanities, Barcelona, Spain
  - João Cascalheira:
      orcid: https://orcid.org/0000-0003-0321-8892
      institute:
        - Interdisciplinary Center for Archaeology and the Evolution of Human Behaviour, University of Algarve, Faro, Portugal
title-block-published: "Last updated"
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  The Upper Palaeolithic (UP) of westernmost Europe was marked by technological and cultural transformations and abrupt climatic shifts. The Iberian Peninsula, particularly southwestern Iberia, served as a refugia and key eco-cultural niche, making it a key region for UP studies. Vale Boi, the only site in southwestern Iberia with a nearly complete UP sequence, provides critical insights into technological, economical and mobility patterns over time. This study examines lithic raw material procurement and technological organisation at Vale Boi, focusing on the differences between local and non-local cherts. We analysed chert assemblages from Gravettian, Proto-Solutrean, and Solutrean levels, integrating previously published techno-typological and raw material macroscopic and petrographic data. Our results indicate that non-local cherts were intensively reduced. The presence of varied techno-typological classes suggests that non-local cherts were not exclusively transported as finished tools, instead highlighting their role as versatile, transportable raw material volumes. Gravettian occupations exhibited greater reliance on non-local cherts, suggesting short-term occupations and higher mobility, while Proto-Solutrean and Solutrean occupations showed increased dependence on local materials, suggesting long-term occupations. These findings expand our understanding of UP technological adaptations in southwestern Iberia, emphasizing the complex role of lithic resources in mobility, settlement, and social networks.
